Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila

### 2018-09-16 phil_paper/jmp_wjv_v2.tex
lots of good stuff

\begin{comment}

@@

%  Shared connections record high levels of monthly consumption, which can expose them to high prices when marginal prices increase with usage . 
This paper analyzes the welfare impacts of pricing policies in the presence of informal water sharing.  Welfare impacts depend not only on heterogeneity in demand for water across the population, but also on costs associated with forming and maintaining water sharing relationships.  To measure these features, I merge administrative data covering 1.5 million water connections in Manila from 2008 to 2015 with detailed survey data on the number of households and people using each water connection.  With comprehensive geographic coverage of half of Manila and a variety of price variation, these data allow for a rich characterization of water demand.  Additional variation in water source choices identify the relative fixed and marginal costs associated with sharing water with a neighbor.  I find that despite relatively high marginal costs of sharing, shared connections allow the city to save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households at low marginal prices.  At the same time, informal sharing networks act as resale markets for water, undermining the extent to which cities can redistribute surplus through increasing marginal prices.

@@

%Monthly quantity choices across different households and at different prices reveal information about household demand.  Observing multiple quantities choices for the same household allows for a rich characterization of demand heterogeneity at the household-level.  
Evaluating different pricing policies requires measuring demand at the household level as well as costs associated with providing piped water.  For information about these attributes, I developed a partnership with one of the regulated, private providers in Manila who provided both billing records as well as information about the regulatory structure, pricing, and production costs.  Complete billing records for each water connection including the monthly meter reading, the total bill charged, and the precise location for each meter serve as the primary measure of quantity choices.  These data span 2008 to 2015, during which connections grew from 800,000 to 1.5 million water connections, generating nearly 100 million transaction records.  Connections are divided into residential (90\%), semi-business (4\%), commercial (5\%), and industrial (1\%).  This paper focuses on residential and semi-business connections.\footnote{Semi-business connections are largely composed of households that own small roadside stands.  These connections provide a useful source of price variation which is described in Appendix~\ref{appendix:ratechangeelasticity}.}  The monthly meter reading is used as the measure of consumption for each connection.  Meter readings are rounded to the nearest cubic meter, and I limit consideration to meter readings below 120 $\text{m}^{3}$ in order to exclude pipe leaks as well as meter reading errors (readings above 120 $\text{m}^{3}$ compose less than 0.5\% of the sample).

@@

\begin{table}
\centering
\caption{Demand Estimates}\label{table:demandestimates}
\input{tables/step_1_estimates_paper} \\
\vspace{.1cm}
{\footnotesize{Total Observations:  473,114   Total Connections:  10,239} }
\end{table}
To interpret these estimates in economic terms, I compute the elasticity of demand with respect to price at the connection-level.  Given a linear demand structure, the elasticity of demand depends on both the price sensitivity and fixed preferences for water.  As a simplified example, consider the calculation for elasticity under a single marginal price and linear demand specification:
\begin{align*}
\frac{\delta Q}{\delta P} \frac{P}{Q} = \alpha_{C} \frac{P}{\gamma_{C}-\alpha_{C} P}
\end{align*}
According to this expression, increasing the fixed preference for water consumption, $\gamma_C$, while holding constant the price sensitivity, $\alpha_C$, mechanically reduces the elasticity.  Many previous studies estimate heterogeneity in $\gamma_C$ alongside a single $\alpha_C$ parameter for the population (\cite{mcrae2014infrastructure}, \cite{olmstead2009reduced}, \cite{szabo2015value}).  This paper builds on these previous studies by estimating heterogeneity in both terms to allow to for a broader range of possibilities, including that larger users may be relatively more sensitive to prices than smaller users.  To compare estimates in this paper with previous papers under non-linear pricing schedules, this approach follows the same formula for computing elasticities by (1) taking draws of $\epsilon$ and $\eta$, (2) computing predicted demand under prices that are 1\% greater than current prices, and (3) comparing the average percentage change in demand before and after the price change.  For tariff blocks equal to zero, prices are increased from 0 to 1 PhP following \cite{szabo2015value}.

@@

\begin{table}
\centering
\caption{Elasticity Estimates}\label{table:elasticity}
\input{tables/elasticity_estimates} \\
\vspace{.5cm}
{\footnotesize Total Observations:  473,114   Total Connections:  10,239 \\
Standard errors are bootstrapped at the connection level.}
\end{table}

@@

\begin{table}
\centering
\caption{Correlation Estimates}\label{table:correstimates}
\input{tables/corr_table_group}\\
\vspace{.3cm}
\footnotesize{1 indexes correlations with first movers \\ while 2 indexes correlations with second movers}
\end{table}

@@

Table~\ref{table:step2estimates} shows the estimation results for the hassle cost estimation.  The hassle cost is estimated to be \input{tables/mean_hassle}PhP per cubic meter of usage, which is approximately \input{tables/hassle_percentage}\unskip\% of the average tariff.  With only 1,247 connections that experience leaks, the estimation recovers wide standard errors for this estimate, rejecting a hassle cost of zero with 10\% confidence.\footnote{Since Steps 2 and 3 of the estimation rely heavily on estimated preference terms, $\gamma$, from Step 1, Appendix~\ref{appendix:incidentalparametertest} tests for bias due to noisy estimation of these incidental parameters and finds little evidence of significant bias.}  These results suggest that the time-costs or other physical challenges transporting water to share with a neighbor may play an important role in the extent to which households choose to use water from a neighbor's tap.

@@

\begin{table}
\centering
\caption{Hassle Cost Estimates}\label{table:step2estimates}
\input{tables/step_2_estimates} 
\vspace{.2cm} 
\footnotesize{Leaking Connections:  1,247,   Neighboring Connections:  5,239,  Total Obs:  36,662  \\ Standard errors are bootstrapped at the leaking connection-neighborhood level.  \\
Appendix~\ref{appendix:hasslecostbootstrap} includes more details on the bootstrapping procedure.}
\end{table}

@@

Disaggregating connection-level preferences as well as including households using from vendors combine to expand the sample from 5,000 connections to 6,984 households.

@@

\begin{table}
\centering
\caption{Simulated Method of Moments Estimates}\label{table:smmestimates}
\input{tables/step_3_estimates_group}\\
\vspace{.05cm}
\footnotesize{ Standard errors are bootstrapped at the connection level.  Total Households:  6,984  }
\end{table}

Table~\ref{table:smmestimates} presents the estimation results, finding a total estimated monthly fixed cost of \input{tables/fixedfee_group}PhP, which is around 50\% higher than the \input{tables/cfee}PhP monthly fixed fee paid the provider.  This finding suggests that households face sizable barriers to connecting to the water network such as maintenance land tenure insecurity, or administrative costs.  This total fixed fee is also comparable to the total variable price paid to the provider each month in Table~\ref{descriptives}, suggesting that fixed costs compose almost half of the total costs of using piped water.

@@

The treated group on the right-hand side shows a relatively stable pre-trend interrupted by a substantial increase in the number of connections just following the discount program implementation.  This increase lasts for a couple years before returning to pre-program levels.  These trends provide evidence that households who are sharing water or purchasing from vendors are gradually learning about the price change and choosing to purchase their own connections.  After a couple years, this additional demand for connections has been satisfied, letting migration take over as the sole driver of new connections.
\begin{figure}
\centering
\caption{Reduced Form: Connection Fee Discount}\label{figure:discountreducedform}
\includegraphics[scale=.57]{tables/diff_595_untreated.pdf}
\includegraphics[scale=.57]{tables/diff_595_treated.pdf}
\end{figure}

@@

Despite these caveats, the fact that the reduced-form and model-based estimates are similar in magnitude lends support to the counterfactual exercises in Section~\ref{section:counterfactualexercises}.  Table~\ref{table:oosestimates} calculates the impacts of this connection fee discount on water source choices as well as consumer welfare.  Consumer welfare is computed as the difference between current utility and the utility that each household would receive from using vended water.  Since utility is assumed to be quasi-linear, consumer welfare is reported in PhP/month.  Lowering the fixed fee leads to substitution away from both vendor and neighbor water sources, which improves consumer welfare especially for the bottom \input{tables/group_percentile}\unskip\% of users.

@@

\begin{table}
\centering
\caption{Mobile and PAWS Water Source Choices}\label{table:pollfishpawsmeans}
\input{tables/pollfish_paws_means_1} 
\vspace{.3cm}
\footnotesize{The PAWS data have been adjusted to reflect that \\ shared connections serve multiple households.}
\end{table}

@@

\begin{table}
\centering
\caption{Mobile and PAWS Descriptives by Water Sharing Status}\label{table:pollfishpawssharing}
\input{tables/pollfish_paws_diff_1}
\end{table}

@@

@@

\begin{figure}
\caption{Consumption Histogram}\label{figure:consumptionhistogram}
\begin{center}
\includegraphics[scale=.7]{tables/tariff_structure_2.pdf}\\
\footnotesize{Consumption values are truncated at 60 to highlight bunching at kink points.}
\end{center}
\end{figure}

\subsection{Time Series Variation in Tariffs}\label{appendix:tariffvariation}
This figure indicates time series variation in prices as a result of the government regulator raising prices incrementally to ensure production costs are covered.

\begin{center}
\includegraphics[scale=.65]{tables/tariff_time_series.pdf}
\end{center}

@@

\begin{table}
\centering
\caption{Summary Statistics and Comparison of Means for Residential, Semi-Business, and Upgraded Connections}\label{table:semittests}
\input{tables/sem_compare_means1}\\
\footnotesize{Data are from the connection survey.}
\end{table}

@@

	\item \textit{ Connections that experience a leak are representative of the population of connections. }  This assumption is supported by minimal economic differences in means of observables between connections with leaks and without in Table~\ref{table:leakbalance} (although some differences are statistically significant).  
\begin{table}
\centering
\caption{Descriptives and Balance Test for Non-Leaking and Leaking Connections}\label{table:leakbalance}
\input{tables/mean_diff_leak1} \\
\footnotesize{Drink is a dummy for drinking tap water.  Water flow is a dummy for low water pressure. \\ Water storage is a dummy for owning a water storage container.  Std. errors result from a T-test of differences in means.}
\end{table}

	\item  \textit{ The decision to disconnect conditional on experiencing a leak is uncorrelated with water demand. }  Table~\ref{table:leakbalancedc} provides partial support by showing few large differences in characteristics between those that disconnect and those that remain connected other than the size of leak.  This evidence suggests that the disconnection decision is driven primarily by variation in the magnitude of the leak.
\begin{table}
\centering
\caption{Balance Test for Disconnection among Leaking Connections}\label{table:leakbalancedc}
\input{tables/mean_diff_DC1.tex} \\
\footnotesize{Drink is a dummy for drinking tap water.  Water flow is a dummy for low water pressure. \\ Water storage is a dummy for owning a water storage container. Std. errors result from a T-test of differences in means.}
\end{table}

@@

\section{Bootstrapping Procedure for Hassle Cost}\label{appendix:hasslecostbootstrap}

The bootstrapping procedure resamples at the level of full neighborhoods including both leaking connection and the neighboring connections, which treats each ``leak'' as its own cluster.  In a small number of bootstrapped samples ($<5\%$), the estimation procedure for this step converges to extreme outliers; this is because these samples have a relatively high share of households choosing zero monthly consumption, which are difficult values for the model to fit.  In these cases, I adjust the routine to drop zero consumption values ($<2\%$ of all values), which allows the procedure to converge to plausible parameter values.

@@

\begin{table}
\centering
\caption{Simulation Test for Incidental Parameter Bias}\label{table:incidentalparameters}
\input{tables/incidental_parameters}\\
\footnotesize{T refers to months observed for each connections.  The first sample includes half of the time periods per connection (around 35 months), while the second sample includes all months.  Both samples include 5,000 accounts.}
\end{table}

@@

\end{comment}

>>>>

@@

%  Shared connections record high levels of monthly consumption, which can expose them to high prices when marginal prices increase with usage . 
This paper analyzes the welfare impacts of pricing policies in the presence of informal water sharing.  Welfare impacts depend not only on heterogeneity in demand for water across the population, but also on costs associated with forming and maintaining water sharing relationships.  To measure these features, I merge administrative data covering 1.5 million water connections in Manila from 2010 to 2015 with detailed survey data on the number of households and people using each water connection.  With comprehensive geographic coverage of half of Manila and a variety of price variation, these data allow for a rich characterization of water demand.  Additional variation in water source choices identify the relative fixed and marginal costs associated with sharing water with a neighbor.  I find that despite relatively high marginal costs of sharing, shared connections allow the city to save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households at low marginal prices.  At the same time, informal sharing networks act as resale markets for water, undermining the extent to which cities can redistribute surplus through increasing marginal prices.

@@

%Monthly quantity choices across different households and at different prices reveal information about household demand.  Observing multiple quantities choices for the same household allows for a rich characterization of demand heterogeneity at the household-level.  
Evaluating different pricing policies requires measuring demand at the household level as well as costs associated with providing piped water.  For information about these attributes, I developed a partnership with one of the regulated, private providers in Manila who provided both billing records as well as information about the regulatory structure, pricing, and production costs.  Complete billing records for each water connection including the monthly meter reading, the total bill charged, and the precise location for each meter serve as the primary measure of quantity choices.  These data span 2010 to 2015, during which connections grew from 800,000 to 1.5 million water connections, generating nearly 100 million transaction records.  Connections are divided into residential (90\%), semi-business (4\%), commercial (5\%), and industrial (1\%).  This paper focuses on residential and semi-business connections.\footnote{Semi-business connections are largely composed of households that own small roadside stands.  These connections provide a useful source of price variation which is described in Appendix~\ref{appendix:ratechangeelasticity}.}  The monthly meter reading is used as the measure of consumption for each connection.  Meter readings are rounded to the nearest cubic meter, and I limit consideration to meter readings below 120 $\text{m}^{3}$ in order to exclude pipe leaks as well as meter reading errors (readings above 120 $\text{m}^{3}$ compose less than 0.5\% of the sample).

@@

\begin{table}
\centering
\caption{Demand Estimates}\label{table:demandestimates}
\input{tables/step_1_estimates_group} \\
\vspace{.1cm}
{\footnotesize{Total Observations:  349,265   Total Connections:  7,151} }
\end{table}
To interpret these estimates in economic terms, I compute the elasticity of demand with respect to price at the connection-level.  Given a linear demand structure, the elasticity of demand depends on both the price sensitivity and fixed preferences for water.  As a simplified example, consider the calculation for elasticity under a single marginal price and linear demand specification:
\begin{align*}
\frac{\delta Q}{\delta P} \frac{P}{Q} = \alpha_{C} \frac{P}{\gamma_{C}-\alpha_{C} P}
\end{align*}
According to this expression, increasing the fixed preference for water consumption, $\gamma_C$, while holding constant the price sensitivity, $\alpha_C$, mechanically reduces the elasticity.  Many previous studies estimate heterogeneity in $\gamma_C$ alongside a single $\alpha_C$ parameter for the population (\cite{mcrae2014infrastructure}, \cite{olmstead2009reduced}, \cite{szabo2015value}).  This paper builds on these previous studies by estimating heterogeneity in both terms to allow to for a broader range of possibilities, including that larger users may be relatively more sensitive to prices than smaller users.  To compare estimates in this paper with previous papers under non-linear pricing schedules, this approach follows the same formula for computing elasticities by (1) taking draws of $\epsilon$ and $\eta$, (2) computing predicted demand under prices that are 1\% greater than current prices, and (3) comparing the average percentage change in demand before and after the price change.  For tariff blocks equal to zero, prices are increased from 0 to 1 PhP following \cite{szabo2015value}.

@@

\begin{table}
\centering
\caption{Elasticity Estimates}\label{table:elasticity}
\input{tables/elasticity_estimates} \\
\vspace{.5cm}
{\footnotesize Total Observations:  349,265   Total Connections:  7,151 \\
Standard errors are bootstrapped at the connection level.}
\end{table}

@@

\begin{table}
\centering
\caption{Correlation Estimates}\label{table:correstimates}
\input{tables/corr_table_group}\\
%\vspace{.3cm}
%\footnotesize{1 indexes correlations with first movers \\ while 2 indexes correlations with second movers}
\end{table}

@@

Table~\ref{table:step2estimates} shows the estimation results for the hassle cost estimation.  The hassle cost is estimated to be \input{tables/mean_hassle}PhP per cubic meter of usage, which is approximately \input{tables/hassle_percentage}\unskip\% of the average tariff.  With only 548 connections that experience leaks, the estimation recovers wide standard errors for this estimate.\footnote{Since Steps 2 and 3 of the estimation rely heavily on estimated preference terms, $\gamma$, from Step 1, Appendix~\ref{appendix:incidentalparametertest} tests for bias due to noisy estimation of these incidental parameters and finds little evidence of significant bias.}  These results suggest that the time-costs or other physical challenges transporting water to share with a neighbor may play an important role in the extent to which households choose to use water from a neighbor's tap.

%, rejecting a hassle cost of zero with 10\% confidence

@@

\begin{table}
\centering
\caption{Hassle Cost Estimates}\label{table:step2estimates}
\input{tables/step_2_estimates} \\
\vspace{.2cm} 
\footnotesize{Leaking Connections:  548,   Neighboring Connections:  2,151,   Total Obs:  20,299  \\ Standard errors are bootstrapped at the leaking connection-neighborhood level. }
\end{table}

\begin{table}
\centering
\caption{Predicted Hassle Cost Distribution}\label{table:step2histogram}
\includegraphics[scale=.7]{tables/hassle_cost_distribution.pdf} \\
%\vspace{cm}
\footnotesize{Excludes top and bottom 1\% outliers}
\end{table}

%Appendix~\ref{appendix:hasslecostbootstrap} includes more details on the bootstrapping procedure.

@@

Disaggregating connection-level preferences as well as including households using from vendors combine to expand the sample from \input{tables/totalhhsmm_group} connections to  households.

@@

\begin{table}
\centering
\caption{Simulated Method of Moments Estimates}\label{table:smmestimates}
\input{tables/step_3_estimates_group}\\
\vspace{.05cm}
\footnotesize{ Standard errors are bootstrapped at the connection level.  Total Households:  \input{tables/totalhhsmm_group}  }
\end{table}

Table~\ref{table:smmestimates} presents the estimation results, finding a total estimated monthly fixed cost of \input{tables/fixedfee_group}PhP, which is around 50\% higher than the \input{tables/cfee_group}PhP monthly fixed fee paid the provider.  This finding suggests that households face sizable barriers to connecting to the water network such as maintenance land tenure insecurity, or administrative costs.  This total fixed fee is also comparable to the total variable price paid to the provider each month in Table~\ref{descriptives}, suggesting that fixed costs compose almost half of the total costs of using piped water.

@@

The treated group on the right-hand side shows a relatively stable pre-trend interrupted by a substantial increase in the number of connections just following the discount program implementation.  This increase lasts for a couple years before returning to pre-program levels.  These trends provide evidence that households who are sharing water or purchasing from vendors are gradually learning about the price change and choosing to purchase their own connections.  After a couple years, this additional demand for connections has been satisfied, letting migration take over as the sole driver of new connections.
\begin{figure}
\centering
\caption{Reduced Form: Connection Fee Discount}\label{figure:discountreducedform}
\includegraphics[scale=.8]{tables/diff_595.pdf}
%\includegraphics[scale=.57]{tables/diff_595_treated.pdf}
\end{figure}

@@

Despite these caveats, the fact that the reduced-form and model-based estimates are similar in magnitude lends support to the counterfactual exercises in Section~\ref{section:counterfactualexercises}.  Table~\ref{table:oosestimates} calculates the impacts of this connection fee discount on water source choices as well as consumer welfare.  Consumer welfare is computed as the difference between current utility and the utility that each household would receive from using vended water.  Since utility is assumed to be quasi-linear, consumer welfare is reported in PhP/month.  Lowering the fixed fee leads to substitution away from both vendor and neighbor water sources, which improves consumer welfare especially for the bottom \input{tables/group_percentile_group}\unskip\% of users.

@@

\begin{table}
\centering
\caption{Mobile and PAWS Water Source Choices}\label{table:pollfishpawsmeans}
\input{tables/pollfish_paws_means} 
\vspace{.3cm}
\footnotesize{The PAWS data have been adjusted to reflect that \\ shared connections serve multiple households.}
\end{table}

@@

\begin{table}
\centering
\caption{Mobile and PAWS Descriptives by Water Sharing Status}\label{table:pollfishpawssharing}
\input{tables/pollfish_paws_diff}
\end{table}

@@

\section{Descriptive for Estimation Sample}\label{appendix:descriptivesest}

\begin{table}
\caption{Descriptives for Estimation Sample}
\begin{center}
\input{tables/new_descriptives_est_sample}
\end{center}
\end{table}

@@

\begin{figure}
\caption{Consumption Histogram}\label{figure:consumptionhistogram}
\begin{center}
\includegraphics[scale=.7]{tables/consumption_histogram.pdf}\\
\footnotesize{Consumption values are truncated at 60 to highlight bunching at kink points.}
\end{center}
\end{figure}

\subsection{Time Series Variation in Tariffs}\label{appendix:tariffvariation}
This figure indicates time series variation in prices as a result of the government regulator raising prices incrementally to ensure production costs are covered.
\begin{figure}
\caption{Regulated Tariff Changes}\label{figure:regulatedtariffchanges}
\begin{center}
\includegraphics[scale=.7]{tables/tariff_time_series.pdf}
\end{center}
\end{figure}

@@

\begin{table}
\centering
\caption{Summary Statistics and Comparison of Means for Residential, Semi-Business, and Upgraded Connections}\label{table:semittests}
\resizebox{\textwidth}{!}{  
\input{tables/sem_compare_means}
}\\
\footnotesize{Data are from the connection survey.}
\end{table}

@@

	\item \textit{ Connections that experience a leak are representative of the population of connections. }  This assumption is supported by minimal economic differences in means of observables between connections with leaks and without in Table~\ref{table:leakbalance} (although some differences are statistically significant).  
\begin{table}
\centering
\caption{Descriptives and Balance Test for Non-Leaking and Leaking Connections}\label{table:leakbalance}
\input{tables/mean_diff_leak} \\
\footnotesize{Drink is a dummy for drinking tap water.  Water flow is a dummy for low water pressure. \\ Water storage is a dummy for owning a water storage container.  Std. errors result from a T-test of differences in means.}
\end{table}

	\item  \textit{ The decision to disconnect conditional on experiencing a leak is uncorrelated with water demand. }  Table~\ref{table:leakbalancedc} provides partial support by showing few large differences in characteristics between those that disconnect and those that remain connected other than the size of leak.  This evidence suggests that the disconnection decision is driven primarily by variation in the magnitude of the leak.
\begin{table}
\centering
\caption{Balance Test for Disconnection among Leaking Connections}\label{table:leakbalancedc}
\input{tables/mean_diff_DC.tex} \\
\footnotesize{Drink is a dummy for drinking tap water.  Water flow is a dummy for low water pressure. \\ Water storage is a dummy for owning a water storage container. Std. errors result from a T-test of differences in means.}
\end{table}

@@

%\section{Bootstrapping Procedure for Hassle Cost}\label{appendix:hasslecostbootstrap}

%The bootstrapping procedure resamples at the level of full neighborhoods including both leaking connection and the neighboring connections, which treats each ``leak'' as its own cluster.  In a small number of bootstrapped samples ($<5\%$), the estimation procedure for this step converges to extreme outliers; this is because these samples have a relatively high share of households choosing zero monthly consumption, which are difficult values for the model to fit.  In these cases, I adjust the routine to drop zero consumption values ($<2\%$ of all values), which allows the procedure to converge to plausible parameter values.

@@

\centering
\caption{Simulation Test for Incidental Parameter Bias}\label{table:incidentalparameters}
\input{tables/incidental_parameters}\\
\footnotesize{T refers to months observed for each connections.  The first sample includes half of the time periods per connection (around 35 months), while the second sample includes all months.  Both samples include 5,000 accounts.}
\end{table}

@@
